---
title: "Allen Coral Atlas Class Descriptions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allen Coral Atlas Class Descriptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This page provides the full class descriptions and colour schemes used by 
ReefMappeR for benthic habitat and geomorphic zone mapping, following the 
[Allen Coral Atlas](https://allencoralatlas.org) classification system.

Full methodology: [allencoralatlas.org/methods](https://allencoralatlas.org/methods/)

---

## Geomorphic Zones

```{r geomorph_table, echo=FALSE, results='asis'}
geomorph <- data.frame(
  Value = c(11, 12, 13, 14, 15, 16, 21, 22, 23, 24, 25),
  Colour = c("#77d0fc", "#2ca2f9", "#c5a7cb", "#92739d", "#614272",
             "#fbdefb", "#10bda6", "#288471", "#cd6812", "#befbff", "#ffba15"),
  Class = c("Shallow Lagoon", "Deep Lagoon", "Inner Reef Flat",
            "Outer Reef Flat", "Reef Crest", "Terrestrial Reef Flat",
            "Sheltered Reef Slope", "Reef Slope", "Plateau",
            "Back Reef Slope", "Patch Reef"),
  Description = c(
    "A shallow, enclosed or semi-enclosed, sheltered, shallow-bottomed area dominated by sediments.",
    "A sheltered, broad body of water, semi- or fully enclosed by reefs. Variable depth with soft bottom dominated by reef sediment.",
    "A low-energy, sediment-dominated, horizontal to gently sloping platform behind the Outer Reef Flat.",
    "Adjoins the seaward margin of the reef; a planar, broad and shallow platform with strong wave-driven zonation.",
    "A zone marking the boundary between the reef flat and reef slope. Typically shallow with highest wave energy absorption.",
    "A broad, flat area of fringing reef directly connected to land, exposed to freshwater runoff, nutrients and sediments.",
    "A submerged, sloping area extending into deep water, sheltered from prevailing wind or current.",
    "A submerged, sloping area extending from the reef crest toward the shelf edge.",
    "A deeper, submerged reef feature with hard bottom, oriented horizontally or gently sloping seaward.",
    "A complex, inner, gently sloping reef zone behind the Reef Flat, sediment-dominated with coral outcrops.",
    "A small, free-standing coral reef within a lagoon, rising from a sandy area."
  )
)

colour_cells <- sprintf('<td style="background-color:%s; width:40px;">&nbsp;</td>', geomorph$Colour)
rows <- sprintf("<tr><td>%s</td>%s<td>%s</td><td>%s</td></tr>",
                geomorph$Value, colour_cells, geomorph$Class, geomorph$Description)
cat('<table><thead><tr><th>Value</th><th>Colour</th><th>Class</th><th>Description</th></tr></thead><tbody>',
    paste(rows, collapse = "\n"),
    '</tbody></table>')
```

---

## Benthic Habitats

```{r benthic_table, echo=FALSE, results='asis'}
benthic <- data.frame(
  Value = c(11, 12, 13, 14, 15, 18),
  Colour = c("#ffffbe", "#e0d05e", "#b19c3a", "#668438", "#ff6161", "#9bcc4f"),
  Class = c("Sand", "Rubble", "Rock", "Seagrass", "Coral/Algae", "Microalgal Mats"),
  Description = c(
    "A soft substrate dominated primarily by fine, unconsolidated sediments.",
    "A habitat of loose, coarse fragments of broken reef material.",
    "Any exposed surface of hard, bare substrate.",
    "Habitats where seagrass is the dominant biota.",
    "Any hard substrate area where living corals and/or algae are growing.",
    "Visible accumulations of microscopic algae within sandy sediments."
  )
)

colour_cells <- sprintf('<td style="background-color:%s; width:40px;">&nbsp;</td>', benthic$Colour)
rows <- sprintf("<tr><td>%s</td>%s<td>%s</td><td>%s</td></tr>",
                benthic$Value, colour_cells, benthic$Class, benthic$Description)
cat('<table><thead><tr><th>Value</th><th>Colour</th><th>Class</th><th>Description</th></tr></thead><tbody>',
    paste(rows, collapse = "\n"),
    '</tbody></table>')
```

---

## Reference

Allen Coral Atlas (2020). Coral reefs of the world. <https://allencoralatlas.org>

Lyons, M.B. et al. (2020). Mapping the world's coral reefs using a global 
multiscale earth observation framework. *Remote Sensing in Ecology and 
Conservation*, 6(4), 557--568. <https://doi.org/10.1002/rse2.157>
