---
title: "Getting Started with ReefMappeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ReefMappeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ReefMappeR)
library(terra)
```

## Overview

ReefMappeR provides a streamlined workflow for mapping and analysing reef 
habitats at the Great Barrier Reef using satellite remote sensing and open 
marine datasets. This vignette walks through a complete analysis from defining 
a region of interest to detecting temporal change in water quality.

The workflow consists of three stages:

1. **Habitat mapping** — load spatial data, visualise and summarise habitats
2. **Water quality** — estimate chlorophyll-a and suspended sediments from Sentinel-2
3. **Change detection** — compare two time points to detect NDCI change

---

## 1. Define a Region of Interest

`set_roi()` is the entry point of every ReefMappeR workflow. It automatically 
loads and crops all bundled datasets — bathymetry, benthic habitat tiles, 
geomorphic zone tiles, and seagrass polygons — to your area of interest.

You can define the ROI either by a bounding box or a Sentinel-2 image:

```{r set_roi, eval=FALSE}
# Option A: bounding box (xmin, xmax, ymin, ymax)
bbox   <- c(150.56, 150.80, -21.51, -21.20)
result <- set_roi(bbox = bbox)

# Option B: Sentinel-2 image extent
s2     <- terra::rast(system.file("extdata", "Sentinel2_example.tif",
                                   package = "ReefMappeR"))
result <- set_roi(s2 = s2)
```

The result is a named list with four elements: `bathymetry`, `benthic`, 
`geomorph`, and `seagrass`. All subsequent functions accept this list as input.

---

## 2. Visualise Habitat Maps

`plot_habitat_map()` creates a two-panel map with bathymetry as background 
and Allen Coral Atlas benthic habitats and geomorphic zones overlaid:

```{r plot_habitat_map, eval=FALSE}
plot_habitat_map(result)
```

The function automatically aggregates large rasters to prevent memory issues, 
applies focal smoothing to bathymetry, and uses the official Allen Coral Atlas 
colour scheme for habitat classes.

---

## 3. Summarise Habitat Statistics

`habitat_summary()` computes per-class area and depth statistics and produces 
a styled summary table:

```{r habitat_summary, eval=FALSE}
out <- habitat_summary(result)
out$table          # styled gt table
out$stats$benthic  # raw benthic statistics
out$stats$geomorph # raw geomorphic statistics
```

`plot_depth_distribution()` visualises the depth distribution of each habitat 
class as boxplots:

```{r plot_depth, eval=FALSE}
plot_depth_distribution(result)
```

---

## 4. Water Quality from Sentinel-2

`calc_ndci()` computes the Normalized Difference Chlorophyll Index (NDCI) 
from Sentinel-2 bands B4 and B5. NDCI is a proxy for chlorophyll-a 
concentration in coastal waters (Mishra & Mishra, 2012):

```{r ndci, eval=FALSE}
s2   <- terra::rast(system.file("extdata", "Sentinel2_example.tif",
                                 package = "ReefMappeR"))
ndci <- calc_ndci(s2)
```

NDCI values range from -1 to 1. Values above 0.3 indicate high chlorophyll-a, 
potentially associated with eutrophication or algal blooms.

`calculate_water_quality()` estimates Total Suspended Sediments (TSS) from 
Sentinel-2 bands B3, B4 and B8:

```{r tss, eval=FALSE}
tss <- calculate_water_quality(s2)
```

---

## 5. Detect Temporal Change

To detect change in chlorophyll-a between two dates, load two Sentinel-2 
images from the same season and compute NDCI for each:

```{r change, eval=FALSE}
s2_24 <- terra::rast(system.file("extdata", "Sentinel2_2024_example.tif",
                                  package = "ReefMappeR"))
s2_25 <- terra::rast(system.file("extdata", "Sentinel2_example.tif",
                                  package = "ReefMappeR"))

ndci_24 <- calc_ndci(s2_24)
ndci_25 <- calc_ndci(s2_25)
```

`assess_reef_change()` computes the pixel-wise NDCI difference:

```{r assess, eval=FALSE}
change <- assess_reef_change(ndci_24, ndci_25)
```

`compare_habitats_change()` classifies the change into three categories and 
visualises the result. The default threshold of 0.1 can be adjusted:

```{r compare, eval=FALSE}
compare_habitats_change(change)
compare_habitats_change(change, threshold = 0.05)
```

---

## References

Allen Coral Atlas (2020). Coral reefs of the world. https://allencoralatlas.org

Lyons, M.B. et al. (2020). Mapping the world's coral reefs using a global 
multiscale earth observation framework. *Remote Sensing in Ecology and 
Conservation*, 6(4), 557–568.

Mishra, S. & Mishra, D.R. (2012). Normalized difference chlorophyll index: 
A novel model for remote sensing of chlorophyll-a concentration in turbid 
productive waters. *Remote Sensing of Environment*, 117, 394–406.
